---
title: "Quick Start Guide (Algebraic API)"
author: "Ian Moran"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide (Algebraic API)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r include = FALSE}
# Load package files
source("../R/classes.R")
source("../R/primitives.R")
source("../R/diagram.R")
source("../R/algebraic-api.R")
source("../R/semantics.R")
source("../R/visualization.R")
```

## What is openstockflow?

**openstockflow** lets you build mathematical models using stock-flow diagrams—a visual way to represent systems with accumulations (stocks) and rates of change (flows).

Think of it like plumbing:
- **Stocks** are tanks that hold quantities
- **Flows** are pipes that move quantities between tanks
- The package automatically generates differential equations and solves them

This guide demonstrates the **algebraic API**, which uses mathematical notation and operators for model construction.

## 5-Minute Tutorial

### Step 1: Create a Simple Model

Let's model exponential decay: a quantity that decreases over time.

```{r decay}
# Define components
X <- stock("X", initial = 100)

decay <- flow("decay",
  rate = function(inputs, params) {
    params$k * inputs$X
  }
)

# Compose: X flows out via decay
decay_model <- X %+% (X %->% decay %->% NULL)
decay_diagram <- finalize(decay_model, reset_ids = TRUE)

# View the structure
print(decay_diagram)
```

This creates the equation: $\frac{dX}{dt} = -k \cdot X$

### Step 2: Solve It

```{r decay_solve, eval = requireNamespace("deSolve", quietly = TRUE)}
if (requireNamespace("deSolve", quietly = TRUE)) {
  result <- solve_diagram(
    decay_diagram,
    times = seq(0, 10, by = 0.5),
    params = c(k = 0.2)
  )

  # View first few rows
  head(result, 10)
}
```

### Step 3: Visualize It

```{r decay_plot, eval = requireNamespace("deSolve", quietly = TRUE)}
if (requireNamespace("deSolve", quietly = TRUE)) {
  plot_solution(result)
}
```

The analytical solution is $X(t) = X_0 e^{-kt} = 100 e^{-0.2t}$

## Example: Population Growth with Carrying Capacity

Now something more interesting: logistic growth where a population grows but slows as it approaches a maximum capacity.

```{r logistic}
# Define stock
P <- stock("Population", initial = 10)

# Define flow
births <- flow("births",
  rate = function(inputs, params) {
    # Growth slows as population approaches carrying capacity
    growth_rate <- params$r * (1 - inputs$Population / params$K)
    growth_rate * inputs$Population
  }
)

# Compose: births flow into Population
logistic <- (NULL %->% births %->% P) %+% P
logistic_diagram <- finalize(logistic, reset_ids = TRUE)

print(logistic_diagram)
```

This implements: $\frac{dP}{dt} = r \cdot P \cdot (1 - \frac{P}{K})$

```{r logistic_solve, eval = requireNamespace("deSolve", quietly = TRUE)}
if (requireNamespace("deSolve", quietly = TRUE)) {
  logistic_result <- solve_diagram(
    logistic_diagram,
    times = seq(0, 50, by = 0.5),
    params = c(r = 0.5, K = 1000)
  )

  # Plot growth curve
  plot_solution(logistic_result)

  # Check that it approaches carrying capacity
  cat("Final population:", tail(logistic_result$Population, 1), "\n")
  cat("Carrying capacity:", 1000, "\n")
}
```

## Example: Two-Stock System

A predator-prey model (simplified Lotka-Volterra):

```{r predator_prey}
# Define stocks
Prey <- stock("Prey", initial = 100)
Predator <- stock("Predator", initial = 10)

# Define flows
prey_growth <- flow("prey_growth",
  rate = function(inputs, params) {
    params$alpha * inputs$Prey
  }
)

predation <- flow("predation",
  rate = function(inputs, params) {
    params$beta * inputs$Prey * inputs$Predator
  }
)

predator_death <- flow("predator_death",
  rate = function(inputs, params) {
    params$delta * inputs$Predator
  }
)

# Compose the system
predator_prey <- Prey %+% Predator %+%
  (NULL %->% prey_growth %->% Prey) %+%
  (Prey %->% predation %->% Predator) %+%
  (Predator %->% predator_death %->% NULL)

pp_diagram <- finalize(predator_prey, reset_ids = TRUE)

print(pp_diagram)
```

Equations:
- $\frac{d(\text{Prey})}{dt} = \alpha \cdot \text{Prey} - \beta \cdot \text{Prey} \cdot \text{Predator}$
- $\frac{d(\text{Predator})}{dt} = \beta \cdot \text{Prey} \cdot \text{Predator} - \delta \cdot \text{Predator}$

```{r predator_prey_solve, eval = requireNamespace("deSolve", quietly = TRUE)}
if (requireNamespace("deSolve", quietly = TRUE)) {
  pp_result <- solve_diagram(
    pp_diagram,
    times = seq(0, 100, by = 0.1),
    params = c(alpha = 0.1, beta = 0.002, delta = 0.05)
  )

  # Plot both populations
  plot_solution(pp_result)

  # Phase plot shows cyclic dynamics
  phase_plot(pp_result, "Prey", "Predator")
}
```

## Algebraic API Syntax Guide

The algebraic API uses operators to build models:

### Stock Definition
```{r eval = FALSE}
X <- stock("X", initial = 100)
```

### Flow Definition
```{r eval = FALSE}
f <- flow("f", rate = function(inputs, params) { ... })
```

### Connection Operator: `%->%`
Connects stocks and flows:
```{r eval = FALSE}
A %->% flow1 %->% B  # Flow from A to B
NULL %->% flow2 %->% C  # External inflow to C
D %->% flow3 %->% NULL  # Outflow from D
```

### Combination Operator: `%+%`
Combines components:
```{r eval = FALSE}
model <- stock1 %+% stock2 %+%
         (stock1 %->% flow1 %->% stock2)
```

### Finalization
Convert specification to diagram:
```{r eval = FALSE}
diagram <- finalize(model, reset_ids = TRUE)
```

## Example: SIR Epidemic Model

The classic SIR (Susceptible-Infected-Recovered) model:

```{r sir}
# Define stocks
S <- stock("S", initial = 999)
I <- stock("I", initial = 1)
R <- stock("R", initial = 0)

# Define flows
infection <- flow("infection",
  rate = function(inputs, params) {
    params$beta * inputs$S * inputs$I / 1000
  }
)

recovery <- flow("recovery",
  rate = function(inputs, params) {
    params$gamma * inputs$I
  }
)

# Compose model
sir <- S %+% I %+% R %+%
  (S %->% infection %->% I) %+%
  (I %->% recovery %->% R)

sir_diagram <- finalize(sir, reset_ids = TRUE)
print(sir_diagram)
```

This creates:
- $\frac{dS}{dt} = -\beta \cdot S \cdot I / N$
- $\frac{dI}{dt} = \beta \cdot S \cdot I / N - \gamma \cdot I$
- $\frac{dR}{dt} = \gamma \cdot I$

```{r sir_solve, eval = requireNamespace("deSolve", quietly = TRUE)}
if (requireNamespace("deSolve", quietly = TRUE)) {
  sir_result <- solve_diagram(
    sir_diagram,
    times = seq(0, 100, by = 1),
    params = c(beta = 0.5, gamma = 0.1)
  )

  # Plot all compartments
  plot_solution(sir_result)

  # Check conservation: S + I + R = 1000
  total <- sir_result$S + sir_result$I + sir_result$R
  cat("Population conserved:", all(abs(total - 1000) < 1e-6), "\n")
}
```

## Key Concepts Summary

### Stocks
- Represent accumulations (population, water in tank, money in account)
- Have initial values
- Change over time according to flows
- Created with `stock(name, initial = value)`

### Flows
- Represent rates of change
- Connect stocks using the `%->%` operator
- Defined by rate functions that depend on:
  - Current stock values (`inputs$StockName`)
  - Parameters (`params$ParamName`)
- Created with `flow(name, rate = function(inputs, params) { ... })`

### Parameters
- Constants that control model behavior
- Passed to `solve_diagram()`
- Examples: birth rate, infection rate, decay constant

### Rate Functions
Flow rate functions have the signature:
```r
function(inputs, params) {
  # inputs$X gives current value of stock X
  # params$k gives parameter k
  # return a single numeric value (the rate)
}
```

### Model Building Pattern
1. Define stocks with `stock()`
2. Define flows with `flow()`
3. Connect with `%->%` operator
4. Combine with `%+%` operator
5. Finalize with `finalize()`
6. Solve with `solve_diagram()`

## What's Next?

- **Introduction (Algebraic)**: More examples including SEIR model with population conservation
- **Composition (Algebraic)**: Learn to build complex models from simpler components
- **Mathematical Foundations**: Category theory and functorial semantics

## Common Patterns

### Conservation

When modeling conserved quantities, you can use sum variables:

```{r conservation_example, eval = FALSE}
# After creating model with stocks
diagram <- finalize(model, reset_ids = TRUE)
diagram <- diagram %>%
  add_sum_variable("Total", stocks = c("Stock1", "Stock2", "Stock3"))
```

The total will remain constant.

### Feedback Loops

Stocks can influence their own flows:

```{r feedback_example, eval = FALSE}
P <- stock("Population", initial = 100)

growth <- flow("growth",
  rate = function(inputs, params) {
    # Population influences its own growth rate
    params$r * inputs$Population
  }
)

model <- P %+% (NULL %->% growth %->% P)
```

### Coupled Systems

Multiple stocks affecting each other:

```{r coupled_example, eval = FALSE}
A <- stock("A", initial = 100)
B <- stock("B", initial = 50)

A_to_B <- flow("A_to_B",
  rate = function(inputs, params) {
    params$k1 * inputs$A
  }
)

B_to_A <- flow("B_to_A",
  rate = function(inputs, params) {
    params$k2 * inputs$B
  }
)

model <- A %+% B %+%
  (A %->% A_to_B %->% B) %+%
  (B %->% B_to_A %->% A)
```

## Algebraic vs Pipe-Based API

Both APIs create the same underlying models. Choose based on preference:

**Algebraic API** (this guide):
```{r eval = FALSE}
S <- stock("S", 999)
I <- stock("I", 1)
f <- flow("infection", rate = ...)
model <- S %+% I %+% (S %->% f %->% I)
diagram <- finalize(model)
```

**Pipe-Based API** (see quickstart.Rmd):
```{r eval = FALSE}
diagram <- stock_flow_diagram() %>%
  add_stock("S", initial = 999) %>%
  add_stock("I", initial = 1) %>%
  add_flow("infection", from = "S", to = "I", rate = ...)
```

## Tips

1. **Start simple**: Build up complexity gradually
2. **Use meaningful names**: Stock and flow names become equation variable names
3. **Check units**: Make sure rates have correct dimensions
4. **Validate**: Check that your model behaves as expected for extreme parameter values
5. **Visualize first**: Use `plot(diagram)` to see structure before solving

## Getting Help

```{r eval = FALSE}
?stock          # Help on stock creation
?flow           # Help on flow creation
?`%->%`         # Help on connection operator
?finalize       # Help on finalizing models
?solve_diagram  # Help on solving
```

## Session Info

```{r session_info}
sessionInfo()
```
