---
title: "Quick Start Guide"
author: "Ian Moran"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r include = FALSE}
# Load package files
source("../R/classes.R")
source("../R/primitives.R")
source("../R/diagram.R")
source("../R/algebraic-api.R")
source("../R/semantics.R")
source("../R/visualization.R")
library(magrittr)
```

## What is openstockflow?

**openstockflow** lets you build mathematical models using stock-flow diagrams—a visual way to represent systems with accumulations (stocks) and rates of change (flows).

Think of it like plumbing:
- **Stocks** are tanks that hold quantities
- **Flows** are pipes that move quantities between tanks
- The package automatically generates differential equations and solves them

## 5-Minute Tutorial

### Step 1: Create a Simple Model

Let's model exponential decay: a quantity that decreases over time.

```{r decay}
decay_model <- stock_flow_diagram(reset_ids = TRUE) %>%
  add_stock("X", initial = 100) %>%
  add_flow("decay",
    from = "X",
    to = NULL,  # NULL means "out of the system"
    rate = function(inputs, params) {
      params$k * inputs$X
    }
  )

# View the structure
print(decay_model)
```

This creates the equation: $\frac{dX}{dt} = -k \cdot X$

### Step 2: Solve It

```{r decay_solve, eval = requireNamespace("deSolve", quietly = TRUE)}
if (requireNamespace("deSolve", quietly = TRUE)) {
  result <- solve_diagram(
    decay_model,
    times = seq(0, 10, by = 0.5),
    params = c(k = 0.2)
  )

  # View first few rows
  head(result, 10)
}
```

### Step 3: Visualize It

```{r decay_plot, eval = requireNamespace("deSolve", quietly = TRUE)}
if (requireNamespace("deSolve", quietly = TRUE)) {
  plot_solution(result)
}
```

The analytical solution is $X(t) = X_0 e^{-kt} = 100 e^{-0.2t}$

## Example: Population Growth with Carrying Capacity

Now something more interesting: logistic growth where a population grows but slows as it approaches a maximum capacity.

```{r logistic}
logistic <- stock_flow_diagram(reset_ids = TRUE) %>%
  add_stock("Population", initial = 10) %>%
  add_flow("births",
    from = NULL,
    to = "Population",
    rate = function(inputs, params) {
      # Growth slows as population approaches carrying capacity
      growth_rate <- params$r * (1 - inputs$Population / params$K)
      growth_rate * inputs$Population
    }
  )

print(logistic)
```

This implements: $\frac{dP}{dt} = r \cdot P \cdot (1 - \frac{P}{K})$

```{r logistic_solve, eval = requireNamespace("deSolve", quietly = TRUE)}
if (requireNamespace("deSolve", quietly = TRUE)) {
  logistic_result <- solve_diagram(
    logistic,
    times = seq(0, 50, by = 0.5),
    params = c(r = 0.5, K = 1000)
  )

  # Plot growth curve
  plot_solution(logistic_result)

  # Check that it approaches carrying capacity
  cat("Final population:", tail(logistic_result$Population, 1), "\n")
  cat("Carrying capacity:", 1000, "\n")
}
```

## Example: Two-Stock System

A predator-prey model (simplified Lotka-Volterra):

```{r predator_prey}
predator_prey <- stock_flow_diagram(reset_ids = TRUE) %>%
  add_stock("Prey", initial = 100) %>%
  add_stock("Predator", initial = 10) %>%
  add_flow("prey_growth",
    from = NULL,
    to = "Prey",
    rate = function(inputs, params) {
      params$alpha * inputs$Prey
    }
  ) %>%
  add_flow("predation",
    from = "Prey",
    to = "Predator",
    rate = function(inputs, params) {
      params$beta * inputs$Prey * inputs$Predator
    }
  ) %>%
  add_flow("predator_death",
    from = "Predator",
    to = NULL,
    rate = function(inputs, params) {
      params$delta * inputs$Predator
    }
  )

print(predator_prey)
```

Equations:
- $\frac{d(\text{Prey})}{dt} = \alpha \cdot \text{Prey} - \beta \cdot \text{Prey} \cdot \text{Predator}$
- $\frac{d(\text{Predator})}{dt} = \beta \cdot \text{Prey} \cdot \text{Predator} - \delta \cdot \text{Predator}$

```{r predator_prey_solve, eval = requireNamespace("deSolve", quietly = TRUE)}
if (requireNamespace("deSolve", quietly = TRUE)) {
  pp_result <- solve_diagram(
    predator_prey,
    times = seq(0, 100, by = 0.1),
    params = c(alpha = 0.1, beta = 0.002, delta = 0.05)
  )

  # Plot both populations
  plot_solution(pp_result)

  # Phase plot shows cyclic dynamics
  phase_plot(pp_result, "Prey", "Predator")
}
```

## Alternative Syntax: Algebraic API

For those who prefer mathematical notation, there's an algebraic API:

```{r algebraic}
# Define components
S <- stock("S", initial = 999)
I <- stock("I", initial = 1)
R <- stock("R", initial = 0)

infection <- flow("infection",
  rate = function(inputs, params) {
    params$beta * inputs$S * inputs$I / 1000
  }
)

recovery <- flow("recovery",
  rate = function(inputs, params) {
    params$gamma * inputs$I
  }
)

# Compose with operators: %->% connects, %+% combines
sir <- S %+% I %+% R %+%
  (S %->% infection %->% I) %+%
  (I %->% recovery %->% R)

sir_diagram <- finalize(sir, reset_ids = TRUE)
print(sir_diagram)
```

This creates the classic SIR epidemic model!

## Key Concepts Summary

### Stocks
- Represent accumulations (population, water in tank, money in account)
- Have initial values
- Change over time according to flows

### Flows
- Represent rates of change
- Have `from` and `to` stocks (or NULL for external sources/sinks)
- Defined by rate functions that depend on:
  - Current stock values (`inputs$StockName`)
  - Parameters (`params$ParamName`)

### Parameters
- Constants that control model behavior
- Passed to `solve_diagram()`
- Examples: birth rate, infection rate, decay constant

### Rate Functions
Flow rate functions have the signature:
```r
function(inputs, params) {
  # inputs$X gives current value of stock X
  # params$k gives parameter k
  # return a single numeric value (the rate)
}
```

## What's Next?

- **Introduction vignette**: More examples including SEIR model with population conservation
- **Composition**: Learn to build complex models from simpler components (future)
- **Stratification**: Add dimensions like age groups or geographic regions (future)

## Common Patterns

### Conservation

When modeling conserved quantities:

```{r conservation_example, eval = FALSE}
model %>%
  add_sum_variable("Total", stocks = c("Stock1", "Stock2", "Stock3"))
```

The total will remain constant.

### Feedback Loops

Stocks can influence their own flows:

```{r feedback_example, eval = FALSE}
model %>%
  add_stock("Population", initial = 100) %>%
  add_flow("growth",
    from = NULL,
    to = "Population",
    rate = function(inputs, params) {
      # Population influences its own growth rate
      params$r * inputs$Population
    }
  )
```

### Coupled Systems

Multiple stocks affecting each other:

```{r coupled_example, eval = FALSE}
model %>%
  add_stock("A", initial = 100) %>%
  add_stock("B", initial = 50) %>%
  add_flow("A_to_B",
    from = "A",
    to = "B",
    rate = function(inputs, params) {
      params$k1 * inputs$A
    }
  ) %>%
  add_flow("B_to_A",
    from = "B",
    to = "A",
    rate = function(inputs, params) {
      params$k2 * inputs$B
    }
  )
```

## Tips

1. **Start simple**: Build up complexity gradually
2. **Use meaningful names**: Stock and flow names become equation variable names
3. **Check units**: Make sure rates have correct dimensions
4. **Validate**: Check that your model behaves as expected for extreme parameter values
5. **Visualize first**: Use `plot(diagram)` to see structure before solving

## Getting Help

```{r eval = FALSE}
?stock_flow_diagram  # Help on diagram construction
?solve_diagram       # Help on solving
?plot.StockFlowDiagram  # Help on visualization
```

## Session Info

```{r session_info}
sessionInfo()
```
